---
title: "Statistical Learning Project"
subtitle: "Homework 1: Unsupervised Learning"
output:
  html_document: 
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
date: "2024-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# Libraries 
```{r}
library(dplyr)
library(GGally)
library(ggplot2)
library(factoextra)
library(cluster)
library(mclust)
library(kernlab)
library(MASS)
```

# Dataset and variable selection
We have decided to use the R-library NHANES(National Health and Nutrition Examination Survey) dataset for our project. It is a dataset that shows the health data of survey takers in the year intervals 2009-2010 and 2011-2012. This dataset has 10000 observations and 76 variables. 
```{r}
install.packages("NHANES")
library(NHANES)
```
If the previous code doesn't work due to different version of RStudio, the 
following code should work. Please set your working directory to be the folder 
with all the documents of our homework. 
```{r}
NHANES <- read.csv("NHANES_data.csv")
```

```{r}
data("NHANES")
head(NHANES)
```

Looking at the first three observations' ID, we see that there are some repetitions. 
Our first step will be to remove those repetitions. Then looking at the summary of 
the data, we can see that there are columns that have a lot of NA values and variables 
that are factors. For our purposes of using this data for unsupervised learning,
we can remove those variables. 
```{r}
# choose only the unique observations 
unique_data <- NHANES %>% distinct(ID, .keep_all = TRUE)
summary(unique_data)
# Choose which variables will be most useful
my_data = unique_data %>% dplyr::select("Gender", "Age", "HHIncomeMid", "Weight", "Height", "BMI",
                                 "Pulse", "BPSysAve", "BPDiaAve", "TotChol", 
                                 "UrineVol1", "UrineFlow1", "DaysPhysHlthBad", 
                                 "DaysMentHlthBad", "SleepHrsNight", "PhysActiveDays",
                                 "SexAge")
```
# Dealing with NA values
```{r}
# Find which rows have the most NA values
na_count <- rowSums(is.na(my_data))
unique_na_counts <- unique(na_count)
unique_na_counts # there are rows with missing from 0 - 14
# As some rows have a lot of missing data, it would be best if we remove them. 
# Hence any rows that have more than 6 missing values will be removed
new_data <- my_data[!(na_count %in% c(6, 7, 8, 9, 10, 11, 12, 13, 14)), ]
# We are left with 5559 observations, which is good.
summary(new_data) 
# With the summary we can see that every column has a lot of missing values, 
# but PhysActiveDays and SexAge have the most(>2100), so it would be better 
# to remove those columns.
new_data = new_data %>% dplyr::select(-PhysActiveDays, -SexAge)
```
For the rest of the variables we will replace the NA values with the average of each variable.
```{r}
# As BMI is calculated based on weight and height, we can replace the the NA 
# values in that column by calculating it. 
new_data$Weight[is.na(new_data$Weight)] <- mean(new_data$Weight, na.rm = TRUE)
new_data$Height[is.na(new_data$Height)] <- mean(new_data$Height, na.rm = TRUE)
new_data$BMI[is.na(new_data$BMI)] <- new_data$Weight[is.na(new_data$BMI)] / (new_data$Height[is.na(new_data$BMI)]^2)
new_data <- new_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
```
As that function will make all of the integer values into numerical,
we will transform them back to integers.
```{r}
new_data$Pulse = as.integer(new_data$Pulse)
new_data$BPSysAve = as.integer(new_data$BPSysAve)
new_data$BPDiaAve = as.integer(new_data$BPDiaAve)
new_data$UrineVol1 = as.integer(new_data$UrineVol1)
new_data$DaysPhysHlthBad = as.integer(new_data$DaysPhysHlthBad)
new_data$DaysMentHlthBad = as.integer(new_data$DaysMentHlthBad)
new_data$SleepHrsNight = as.integer(new_data$SleepHrsNight)
sum(rowSums(is.na(new_data)))  # check if there are any NA values now
```
# Feature engineering. 
Creating new variables that may be of use for us.
```{r}
# Any abnormal BPRatio may indicate cardiac issues 
# As there are some observations with BPDiaAve = 0, which would not work for us,
# we have decided to remove those records. 
new_data <- new_data %>% filter(BPDiaAve != 0)
new_data$BPRatio <- new_data$BPSysAve / new_data$BPDiaAve

# Total Bad Health Days will show us how many days of the year the person didn't feel good.
new_data$TotalBadHealthDays <- new_data$DaysPhysHlthBad + new_data$DaysMentHlthBad

# SleepAdjusted shows if based on age, a person is sleeping enough. We will remove
# SleepHrsNight and replace it with this variable as it would give a more accurate 
# interpretation if the person sleeps enough for their age.
new_data$SleepAdjusted <- ifelse(new_data$Age < 18, new_data$SleepHrsNight / 9, 
                           ifelse(new_data$Age < 65, new_data$SleepHrsNight / 7, new_data$SleepHrsNight / 7))
new_data = new_data %>% dplyr::select(-SleepHrsNight)

# Cholesterol BMI Ration shows the relation between the cholesterol levels and body mass.
new_data$CholesterolBMIRatio <- new_data$TotChol/new_data$BMI

# PulseAgeRatio can indicate health problems if the ratio is abnormal for the age 
new_data$PulseAgeRatio <- new_data$Pulse/new_data$Age
```

# Normalization of variables:
```{r}
normalize_z <- function(x) {(x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)}
normalized_new_data <- new_data %>%
  mutate(across(where(is.numeric), normalize_z))
numerical_data = normalized_new_data %>% dplyr::select(where(is.numeric))
summary(numerical_data)
```

# Dealing with outliers:
```{r}
boxplot(numerical_data[, sapply(numerical_data, is.numeric)],las = 2, cex.axis = 0.5)
```
```{r}
# 1) Identify them
identify_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  outliers <- x < (Q1 - 1.5 * IQR_value) | x > (Q3 + 1.5 * IQR_value)
  return(outliers)
}

# Apply outlier function to each numeric column
outlier_matrix <- sapply(numerical_data[, sapply(numerical_data, is.numeric)], identify_outliers)

# Check which columns and rows have outliers
outliers_by_variable <- colSums(outlier_matrix) # Number of outliers per variable
outliers_by_variable
```

From that we can see that we have many outliers in the columns DaysPhysHlthBad,
DaysMentHlthBad, and TotalBadHealthDays. From this we can judge that they have high 
uniqueness relative to the rest of the variables, which can mess with our 
models, so it would be best to remove these columns and do the unsupervised
learning on the rest of the variables. 
```{r}
numerical_data = numerical_data %>% dplyr::select(-DaysPhysHlthBad, -DaysMentHlthBad, 
                                           -TotalBadHealthDays)
```

Before removing the outliers, we can check the correlation matrix to see if we 
can remove some columns that don't show any strong relationships or some that 
have only one strong relation.
```{r}
ggcorr(numerical_data)
```

From this we can see that CholestrolBMIRatio only has weak relations, 
UrineVol1 and UrineFlow1 are strongly corelated, so we can remove one of them
(remove UrineFlow1 as it has more outliers). We decided to keep both BMI and
Weight for now. 
```{r}
numerical_data = numerical_data %>% dplyr::select(-CholesterolBMIRatio, -UrineFlow1)
```

2) Remove rows with outliers
```{r}
# new outlier matrix
outlier_matrix <- sapply(numerical_data[, sapply(numerical_data, is.numeric)], identify_outliers)
cleaned_data <- numerical_data[!rowSums(outlier_matrix), ]
```
Checking the standard deviation
```{r}
sd_values = sapply(cleaned_data[, sapply(cleaned_data, is.numeric)], sd, na.rm = TRUE)
sd_df <- data.frame(Variable = names(sd_values), SD = sd_values)
ggplot(sd_df, aes(x = reorder(Variable, -SD), y = SD)) +
  geom_bar(stat = "identity", fill = "pink") +
  coord_flip() +  # Flip the chart for easier readability
  theme_minimal() +
  labs(title = "Standard Deviation of Each Variable", x = "Variable", y = "Standard Deviation")
```

From this we can see that BPRatio has a relatively smaller standard deviation,
which can mean that it does not contribute much to the variability of the data.
Hence, we can remove it. 
```{r}
numerical_data = numerical_data %>% dplyr::select(-BPRatio)
outlier_matrix <- sapply(numerical_data[, sapply(numerical_data, is.numeric)], identify_outliers)
final_cleaned_data <- numerical_data[!rowSums(outlier_matrix), ]
boxplot(final_cleaned_data[, sapply(final_cleaned_data, is.numeric)],las = 2, cex.axis = 0.5)
```
# Visualization

1. Create histograms to show the distribution of each variable.

This allows us to see which variables are normally distributed, which are skewed, 
which are multimodal(have peaks in the histogram), or which have low variance.
```{r}
numeric_vars <- c("Age", "HHIncomeMid", "Weight", "Height", "BMI", "Pulse", "BPSysAve", 
                  "BPDiaAve", "TotChol", "UrineVol1", "SleepAdjusted", "PulseAgeRatio")
for (var in numeric_vars) {
  p <- ggplot(final_cleaned_data, aes_string(var)) +
    geom_histogram(bins = 30, fill = "pink", color = "red") +
    theme_minimal() +
    ggtitle(paste("Distribution of", var)) +
    xlab(var) + ylab("Frequency") +
    theme(plot.title = element_text(hjust = 0.5))
  
  print(p)
}
```

If we were to use UrineVol1 or PulseAgeRatio for clustering, we will need to 
transform them as they are skewed. SleepAdjusted has some peaks, which suggest 
that it already has some natural 
clusters. 

# Deal with skewness. 

Only UrineVol1 and PulseAgeRatio really exhibits heavy skewness. 
```{r}
# for UrineVol1
offset <- abs(min(final_cleaned_data$UrineVol1)) + 1  
boxcox_result <- boxcox((final_cleaned_data$UrineVol1 + offset) ~ 1)
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
final_cleaned_data$UrineVol1 <- ((final_cleaned_data$UrineVol1 + offset)^lambda - 1) / lambda

p <- ggplot(final_cleaned_data, aes_string("UrineVol1")) +
    geom_histogram(bins = 30, fill = "pink", color = "red") +
    theme_minimal() +
    ggtitle(paste("Distribution of", "UrineVol1")) +
    xlab("UrineVol1") + ylab("Frequency") +
    theme(plot.title = element_text(hjust = 0.5))
  
print(p)

#for PulseAgeRatio
offset <- abs(min(final_cleaned_data$PulseAgeRatio)) + 1  
boxcox_result <- boxcox((final_cleaned_data$PulseAgeRatio + offset) ~ 1)
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
final_cleaned_data$PulseAgeRatio <- ((final_cleaned_data$PulseAgeRatio + offset)^lambda - 1) / lambda

p <- ggplot(final_cleaned_data, aes_string("PulseAgeRatio")) +
    geom_histogram(bins = 30, fill = "pink", color = "red") +
    theme_minimal() +
    ggtitle(paste("Distribution of", "PulseAgeRatio")) +
    xlab("PulseAgeRatio") + ylab("Frequency") +
    theme(plot.title = element_text(hjust = 0.5))
  
print(p)
```

# Correlation

Correlation of the numerical data we have decided to keep: 
```{r}
ggcorr(final_cleaned_data)
```

From this graph we can see that there aren't a lot of variables that have a high 
correlation between each other. Weight and Height have a strong positive relation 
as well as BPSysAve and BPDiaAve. Weight and BMI have a moderately strong positive 
relation. Variables like UrineVol1 and SleepAdjusted very weak relations with the 
rest. Also, PulseAgeRation shows negative relations with some of the variables.


# PCA
```{r}
pca = prcomp(final_cleaned_data, scale=T)
summary(pca) 
```

We can observe that with 7 variables we are able to explain 84.1% of the data.

```{r}
screeplot(pca,main="Screeplot",col="pink",type="barplot",pch=19) #Variable importance

```

A nicer view than screeplot is Elbow Method (factoextra)

```{r}
fviz_screeplot(pca, addlabels = TRUE) 
```

We can reaffirm with this graph that PCA can be developed correctly with 7 variables 


Let us go component by component

First component: 
```{r}
barplot(pca$rotation[,1], las=2, col="pink")
# This barplot shows all the values in the first column (var 1) of the vector matrix

sum(pca$rotation[,1]^2) #This sum should always be equal to 1

fviz_contrib(pca, choice = "var", axes = 1)
```

We can observe that the most relevant variables for the first component are:
PulseAgeRatio, Age, BPSysAve, Weight and BMI 

The first component appears to represent a general health index, particularly 
cardiovascular health and factors related to age and body weight.


Top 10 people in general health index:

```{r}
# use the data before normalization to get the real values
normal_cleaned_data = new_data %>% dplyr::select(-DaysPhysHlthBad, -DaysMentHlthBad, 
                                      -TotalBadHealthDays, -CholesterolBMIRatio,
                                      -UrineFlow1, -BPRatio)
normal_cleaned_data <- normal_cleaned_data[!rowSums(outlier_matrix), ]

# Get the indices of the top 10 observations along the first principal component
top_indices1 <- order(pca$x[,1], decreasing = TRUE)[1:10]

# Use these indices to subset the original dataset and view the variable information
top_observations1 <- normal_cleaned_data[top_indices1, ]

# Display the top observations
print(top_observations1)
```

Second component:

```{r}
barplot(pca$rotation[,2], las=2, col="pink")
barplot(abs(pca$rotation[,2]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 2)
```

We can observe that the most relevant variables for the second component are:
Weight, Height, BMI, Age and PulseAgeRatio

The second component seems to capture a body composition index, focusing on 
characteristics that define physical development, body size, and possibly the 
influence of age on these metrics.


Third component:
```{r}
barplot(pca$rotation[,3], las=2, col="pink")
barplot(abs(pca$rotation[,3]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 3)
```

We can observe that the most relevant variables for the third component are:
Pulse, HHIncomeMide, Height, BMI and UrineVol1

The third component seems to represent a socioeconomic and physiological health index.


Top 10 people in socioeconomic and physiological health index:

```{r}
# Get the indices of the top 10 observations along the first principal component
top_indices2 <- order(pca$x[,2], decreasing = TRUE)[1:10]

# Use these indices to subset the original dataset and view the variable information
top_observations2 <- normal_cleaned_data[top_indices2, ]

# Display the top observations
print(top_observations2)
```



Fourth component:
```{r}
barplot(pca$rotation[,4], las=2, col="pink")
barplot(abs(pca$rotation[,4]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 4)
```

We can observe that the most relevant variables for the fourth component are:
BPDiaAVe, Pulse, TotChol and BMI

The fourth component appears to represent a cardiovascular and metabolic health risk index.


Fifth component:

```{r}
barplot(pca$rotation[,5], las=2, col="pink")
barplot(abs(pca$rotation[,5]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 5)
```

We can observe that the most relevant variables for the fifth component are:
SleepAdjusted, HHIncomeMid and BPSysAve 

The fifth component seems to represent a lifestyle and socioeconomic health index.



Sixth component:

```{r}
barplot(pca$rotation[,6], las=2, col="pink")
barplot(abs(pca$rotation[,6]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 6)
```

We can observe that the most relevant variables for the sixth component are:
UrineVol1, SleepAdjusted and HHIncomeMid  

PC6 may represent a component associated with lifestyle and health factors 
influenced by socioeconomic status, sleep patterns, and physical health indicators.



Seventh component:

```{r}
barplot(pca$rotation[,7], las=2, col="pink")
barplot(abs(pca$rotation[,7]), las=2, col="pink")

fviz_contrib(pca, choice = "var", axes = 7)
```

We can observe that the most relevant variables for the seventh component are:
TotChol, UrineVol1, Height, BPSysAve and SleepAdjusted   

PC7 might capture aspects of cardiovascular and metabolic health, influenced by 
both physiological markers (like cholesterol and blood pressure) and lifestyle 
elements (like sleep). This component could help differentiate individuals based 
on combined cardiovascular risk factors and body-related attributes.



# Factor analysis

The more factors that we have the better the variance explained will be, but this
can lead to overfitting. Hence, we need to find the balance between the two. 
```{r}
# Factor analysis with all numeric data(should limit them)
fa_result <- factanal(final_cleaned_data, factors = 4, rotation="varimax", scores="regression", lower = 0.01)
fa_result
```
For using all the variables available, we can see that the best solution will be
using four factors as we will get a relatively good variance explained and not be 
in great risk of overfitting. However, as we can see by the chi square statistics, 
this model does not have the best fit. Changing the rotation or the scores here 
doesn't really affect the model. We can now try and remove some of the variables 
that are not well explained by the model. To do that we need to check the uniqueness
of this factor analysis.
```{r}
fa_result$uniquenesses
```
Judging by the uniqueness of the variables, we can see that HHIncomeMid, TotChol, 
UrineVol1 and SleepAdjusted all have above 0.91 uniqueness, which means that more
91% of the variance in these variable cannot be explained by the model. As we 
cannot find better models by changing the factors, rotation or scores, we can 
remove those variables. 
```{r}
final_cleaned_data1 = final_cleaned_data %>% dplyr::select(-HHIncomeMid, -TotChol,
                                                    -UrineVol1, -SleepAdjusted) 
fa_result_1_f4 <- factanal(final_cleaned_data1, factors = 4, rotation="promax", scores="regression",
                        lower = 0.01)
fa_result_1_f4
```
```{r}
fa_result_1_f3 <- factanal(final_cleaned_data1, factors = 3, rotation="promax", scores="regression",
                        lower = 0.01)
fa_result_1_f3
```
For this variables, the best model would be with four factors as well as it explains 
a lot of the data(78.6%) and the risk of overfitting is not very high as its complexity
is not very high. Models will less factors will be less prone to overfitting, but 
their complexity will also decline a lot. 
The fit is still bad so we can try to remove other variables and see if we can find 
a better model. 
Based on uniqueness we can try to remove Pulse as well.
```{r}
final_cleaned_data2 = final_cleaned_data1 %>% dplyr::select(-Pulse)
fa_result_2_f2 <- factanal(final_cleaned_data2, factors = 2, rotation="varimax", scores="Bartlett", lower = 0.01)
fa_result_2_f2
```
```{r}
fa_result_2_f3 <- factanal(final_cleaned_data2, factors = 3, rotation="varimax", scores="Bartlett", lower = 0.01)
fa_result_2_f3
```
The model with 2 factors will explain less than 50% and based on the uniqueness, 
we can see that age, BPSysAve and BPDiaAve are not explained almost at all.
For these variables the best model is with 3 factors as it explains a lot of the
variance(73.8%) and it has only three factors, which means that the risk of 
overfitting is not very high. 
This is our best model for this data as the chi square statistics is very high. 
Based on uniqueness we can try to remove BPDiaAve as well.
```{r}
final_cleaned_data3 = final_cleaned_data2 %>% dplyr::select(-BPDiaAve)
fa_result_3 <- factanal(final_cleaned_data3, factors = 3, rotation="varimax", scores="Bartlett", lower = 0.01)
fa_result_3
```
The best model here will be again using 3 factors as it will explain 85.1% of the 
data, and we finally have a model that has a good fit. Using two factors will not 
capture enough of the underlining patterns and be a worse model. 
We can try removing BPSysAve instead of BPDiaAve now. 
```{r}
final_cleaned_data4 = final_cleaned_data1 %>% dplyr::select(-Pulse, -BPSysAve)
fa_result_4 <- factanal(final_cleaned_data4, factors = 3, rotation="varimax", scores="Bartlett", lower = 0.01)
fa_result_4
```
Here we get a little worse fit than the previous one judjing by the cumulative 
variance and fit. Using two factors will not capture enough of the underlining 
patterns(variance 52%) and be a worse model. 
Lastly, we check without Pulse and PulseAgeRatio:
```{r}
final_cleaned_data5 = final_cleaned_data1 %>% dplyr::select(-Pulse, -PulseAgeRatio)
fa_result_5 <- factanal(final_cleaned_data5, factors = 3, rotation="varimax", scores="Bartlett", lower = 0.01)
fa_result_5
```
This model is fitting okay, but it has two variables with high uniqueness.

# Choosing the best model and explaining its loadings:

Best model overall would be fa_result_3 as it explains the most variance 85.1% 
and has the lowest uniqueness, which means that it captures the underlining patterns
the best. However, it possesses the risk of overfitting. 

The best model for predicting new data would be model fa_result_2_f3 as it is a 
good balance between variance (73.8%), fit, and complexity. The risk of overfitting
is less than with fa_result_3, but it still explains more than fa_result and 
fa_result_1_f4. 

For fa_result_3(best overall model) we have the following uniqueness and loadings:
```{r}
fa_result_3$uniquenesses
```
The model explains the variance in the variables Age, Weight, Height, BMI, and 
PulseAgeRatio very well. BPSysAve is not well explained judging by the high value
of uniqueness. 

The loadings of this model are ...
```{r}
fa_result_3$loadings
```
Factor 1 - explains 35.6% variance
- Age has a high loading(0.990) => F1 is strongly associated with age.
- BPSysAve has a moderate loading (0.454) => BPSysAve has some association with F1
- PulseAgeRatio has a strong negative loading (-0.950) => as F1 increases, 
PulseAgeRatio decreases

Interpretation of F1: Factor 1 captures the variables, which are related to age 
and health indicators that are affected by age, like BPSysAve and PulseAgeRatio. 
This factor could represent the health of a person related to their age.

Factor 2 - explains 30.1% variance
- Weight has a high loading (0.901) => F2 is strongly associated with weight.
- BMI has a high loading (0.977), F2 is also strongly associated with body 
composition and mass.

Interpretation of F2: Factor 2 appears to represent the body mass or body 
composition of a person based on the high loadings of the Weight and BMI variables.

Factor 3 - explains 19.4% variance
- Height has a high loading (0.991) => F3 is strongly associated with height.
- Weight has a moderate loading (0.405) => Weight has some association with F3

Interpretation of F3: Factor 3 is probably a "Height" factor that captures the 
variance in height with minor contribution of weight.


For fa_result_2_f3(best for predicting new data) we have the following uniqueness
and loadings
```{r}
fa_result_2_f3$uniquenesses
```
The model explains the variance in the variables Age, Weight, Height, BMI, and 
PulseAgeRatio very well. BPSysAve and BPDiaAve are not well explained judging by
the high value of uniqueness. 
```{r}
fa_result_2_f3$loadings
```
Both model appear to have similar factors 
- Factor 1 is about the health related to the age. Represents 30.5% variance.
- Factor 2 represent the body mass or body composition of a person. 
  Represents 26.4% variance
- Factor 3 is about the height of the person with a small contribution of the 
  weight. Represents 16.9% variance.
  

# Clustering 

Explore in what groups can people be divided based on their health metrics to 
judge their health risk.
```{r}
clustering_data_health_risk = final_cleaned_data %>% dplyr::select("BMI","BPSysAve",
                                                                  "BPDiaAve", "TotChol",
                                                                  "Pulse")
ggcorr(clustering_data_health_risk)
```


```{r}
heatmap(cor(clustering_data_health_risk),symm=T) 
```
# K means
```{r}
k = 4
set.seed(1)
fit <- kmeans(clustering_data_health_risk, k, nstart = 25, iter.max = 100)
groups = fit$cluster
barplot(table(groups), col="pink")
```

```{r}
fviz_cluster(fit, data = clustering_data_health_risk, geom = c("point"),ellipse.type = 'norm', pointsize=1)+
  theme_minimal()+scale_fill_brewer(palette="Paired")
```

Based on the kmeans clustering plot we can see that there is quite a lot of 
overlapping between the different clusters. We can now check the silhouette width to 
further determine the goodness of the clustering algorithm. 
```{r}
d <- dist(clustering_data_health_risk, method="euclidean")  
sil = silhouette(groups, d)
plot(sil, col=1:5, main="", border=NA)
summary(sil)
```

Values are closer to zero which indicates that the clusters may overlap. The 
negative minimum silhouette width also shows that there are some points that 
have been miscalculated. 

We will now try to find what would be the ideal number of clusters using different
approaches. 
```{r}
fviz_nbclust(clustering_data_health_risk, kmeans, method = 'silhouette', k.max = 20, nstart = 25, iter.max = 100)
```

Here we can see that the optimal number of clusters will be 2. 
```{r}
fviz_nbclust(clustering_data_health_risk, kmeans, method = 'wss', k.max = 20, 
             nstart = 25, iter.max = 100)
```

Judging by the Elbow Plot, we can assume that 4 or 5 cluster will be enough. 
```{r}
fviz_nbclust(clustering_data_health_risk, kmeans, method = 'gap_stat', k.max = 6, nstart = 15, nboot = 50, iter.max = 50)
# Please ignore any warnings shown. It takes a lot of time to be run. 
```

Based on the Gap Statistic Plot, 2 clusters would be the optimal number for our dataset. 

The three graphs give us slightly different answers so we can try and do clustering 
with 2 centers and with 3 centers and after that decide which is better.
```{r}
k = 3
fit_3 <- kmeans(clustering_data_health_risk, k, nstart = 25, iter.max = 100)
groups_3 = fit_3$cluster
fviz_cluster(fit_3, data = clustering_data_health_risk, geom = c("point"),ellipse.type = 'norm', pointsize=1)+
  theme_minimal()+scale_fill_brewer(palette="Paired")
```
```{r}
centers_3=fit_3$centers
for(i in 1:3){
  b = barplot(centers_3[i,], las=2, col="pink")
}
```
```{r}
print(fit_3$size)
print(fit_3$centers)  
```

Based on the centers we can interpret what each cluster will represent.
Cluster 1 - 1256 individuals
- BMI: 0.388 — higher-than-average BMI => could mean higher body weight
- BPSysAve: 0.103 — close to the average
- BPDiaAve: 0.438 — higher than average, but not by a lot 
- TotChol: 0.189 — slightly above average; could increase cardiovascular risk
- Pulse: 0.793 — far above average; the group might be under more physical or 
  metabolic stress, or may worse cardiovascular fitness

Interpretation of C1: This cluster show a group that has higher BMI and pulse rate 
and close to average blood pressure and cholesterol levels. These individuals 
could have an average rick of cardiovascular issues. 

Cluster 2 - 1718 people:
- BMI: -0.376 — people with lower-than-average BMI
- BPSysAve: -0.584 — lower than average(but not severly so), which 
  might mean that this group has healthier blood pressure levels
- BPDiaAve: -0.487 — lower than average
- TotChol: -0.569 — lower cholesterol levels, which is healthier
- Pulse: -0.289 — The pulse rate is also slightly lower 
  than average, which along with lower BMI, TotChol and blood pressure could mean
  better cardiovascular fitness

Interpretation of C2: This cluster shows healthier individuals with below average
BMI, blood pressure, cholesterol and pulse. This could indicate that they have a
lower risk of cardiovascular issues.

Cluster 3 - 1398 individuals
- BMI: 0.099 — close to the average, normal body weight
- BPSysAve: 0.479 — higher-than-average, could increase cardiovascular risk
- BPDiaAve: 0.422 — higher-than-average
- TotChol: 0.489 — higher levels, could increase cardiovascular risk
- Pulse: -0.751 — lower than average; could mean better cardiovascular fitness or 
  it could mean that the body has a compensatory response to manage higher blood pressure
  
Interpretation of C3: This cluster shows individuals with normal BMI, but higher 
blood pressure and cholesterol levels. They might have a higher risk for 
cardiovascular conditions and could benefit from changing their diet and contacting
a doctor to find a way to regulate their blood pressure.

```{r}
d <- dist(clustering_data_health_risk, method="euclidean")  
sil = silhouette(groups_3, d)
plot(sil, col=1:5, main="", border=NA)
summary(sil)
```

Values are close to zero which indicates that the clusters may overlap. The 
negative minimum silhouette width also shows that there are some points that 
have been miscalculated. 
```{r}
k = 2
fit_2 <- kmeans(clustering_data_health_risk, k, nstart = 25, iter.max = 100)
groups_2 = fit_2$cluster
fviz_cluster(fit_2, data = clustering_data_health_risk, geom = c("point"),ellipse.type = 'norm', pointsize=1)+
  theme_minimal()+scale_fill_brewer(palette="Paired")
```
```{r}
print(fit_2$size)
print(fit_2$centers)
```

Based on the centers we can interpret what each cluster will represent.
Cluster 1 - 2194 individuals 
- BMI: - 0.296 - little below the average
- BPSysAve: -0.520 — below average, relatively healthier blood pressure levels.
- BPDiaAve: -0.398 — below average.
- TotChol: -0.475 — lower; could mean a lower risk of cardiovascular issues.
- Pulse: -0.219 — little below average; could mean good cardiovascular health

Interpretation of C1: This cluster shows people that are healthier and have below
average BMI, blood pressure, cholesterol levels and pulse. They could have lower
risk for cardiovascular issues. 

Cluster 2 - 2178 individuals
- BMI: 0.288 — higher BMI; could mean greater body weight
- BPSysAve: 0.430 — higher than average, could mean higher chance cardiovascular risk
- BPDiaAve: 0.540 — above average, supporting the potential for cardiovascular risk
- TotChol: 0.453 — higher-than-average; increased risk for heart diseases
- Pulse: -0.033 — close to average

Interpretation of C2: This cluster shows people that have higher than average BMI, 
blood pressure, cholesterol levels, and pulse. This could indicate a higher chance 
of cardiovascular issues. They would benefit from changing their lifestyle and 
contacting a doctor. 

```{r}
d <- dist(clustering_data_health_risk, method="euclidean")  
sil = silhouette(groups_2, d)
plot(sil, col=1:5, main="", border=NA)
summary(sil)
```

Values are again close to zero, but they are better than for 3 clusters. From 
this we can assume that the clustering with only 2 centers is better than the 
one with 3. 


# PAM 
```{r}
fit_pam <- eclust(clustering_data_health_risk, "pam", stand=TRUE, k=4, graph=F)

fviz_cluster(fit_pam, data = clustering_data_health_risk, geom = c("point"), pointsize=1)+
  theme_minimal()+scale_fill_brewer(palette="Paired")
```

Number of groups by pam:

```{r}
fviz_nbclust(clustering_data_health_risk, pam, method = 'silhouette', k.max = 10)
# This takes a minute to run!
```

Based on this graph, the optimal number of clusters is 2.  

```{r}
fit_pam_2 <- eclust(clustering_data_health_risk, "pam", stand=TRUE, k=2, graph=F)
adjustedRandIndex(fit_2$cluster, fit_pam_2$clustering) 
```
The value(0.79) is close to 1, so we can say that the two classifications are in
agreement. 

```{r}
sil_pam <- silhouette(fit_pam_2$clustering, dist(clustering_data_health_risk))
summary(sil_pam)
```
Worse silhouette widths than k means. 

# Kernel k-means
  
```{r}
fit_ker <- kkmeans(as.matrix(clustering_data_health_risk), centers=2, kernel="rbfdot")
#Also takes a minute to run!
object_ker = list(data = clustering_data_health_risk, cluster = fit_ker@.Data)
fviz_cluster(object_ker, geom = c("point"), ellipse=F,pointsize=1)+
  theme_minimal()+scale_fill_brewer(palette="Paired")
```
```{r}
adjustedRandIndex(fit_2$cluster, fit_ker@.Data) 
```
These are less of in agreement compared to the previous ones. 
```{r}
clusters_ker <- fit_ker@.Data
kernel_matrix <- kernelMatrix(rbfdot(), as.matrix(clustering_data_health_risk))
sil_ker <- silhouette(clusters_ker, as.dist(1 - kernel_matrix))
summary(sil_ker)
```
Very low score, indicating bad classification. 

# EM clustering

```{r}
res.Mclust <- Mclust(clustering_data_health_risk)
summary(res.Mclust)
```

```{r}
fviz_mclust(object = res.Mclust, what = "BIC", pallete = "jco") +
  scale_x_discrete(limits = c(1:10))
```

The optimal number of cluster is 7 here.

```{r}
fviz_mclust(object = res.Mclust, what = "classification", geom = "point",
            pallete = "jco")

```

```{r}
fit_7 <- kmeans(clustering_data_health_risk, 7, nstart = 25, iter.max = 100)
print(adjustedRandIndex(res.Mclust$classification, fit_7$cluster))
```
As we can see by the low values they are not in agreement. 
```{r}
sil_em <- silhouette(res.Mclust$classification, dist(clustering_data_health_risk))
summary(sil_em)
```

The silhoutte widths of the EM clustering are also very low, indicating bad
classification.
